Reducing the Memory Cost of Training Convolutional Neural Networks by CPU Offloading

### 2019-07-09 article.tex
Update figures and full text

\usepackage{amsmath}
 \usepackage{titling}
 \usepackage[ruled, vlined]{algorithm2e}

@@

In recent years Convolutional Neural Networks (CNN) have enabled 
unprecedented progress on a wide range of computer vision tasks. 
However, training large CNNs is a resource intensive task that requires 
specialized Graphical Processing Units (GPU) and highly optimized 
hardware-specific implementations for efficient computation. 
%Given the ubiquity of CNN for practical computer vision applications,
%considerable efforts have been undertaken to optimize their resource consumption.
In training CNNs on modern systems, GPU memory is a major bottleneck limiting both 
the size of the input and the model architectures being considered:
The backpropagation algorithm requires to accumulate the activation 
values of hidden layers in live GPU memory during the forward pass
as these activations are needed for the computation of the weights 
gradient during the backward pass of the backpropagation algorithm.
In this paper, we propose to alleviate this memory bottleneck by 
by offloading the input activation of hidden layers to the CPU:
during the forward pass, we transfer input activations to the CPU upon
computation to free up GPU memory, and transfer these
activations back to the GPU when needed by the backward pass.
On a typical work station with a NVIDIA GTX 2080, we are able reduce the 
memory consumption of traning VGG by xxx\% with a minimal additional 
wall time overhead of xxx\%. 
Our method is orthogonal to other techniques for memory reduction such as 
quantization and sparsification so that it can be easily combined.

@@

\section{Introduction}
% 1. Catchy Intro
Over the last few years, Convolutional Neural Networks (CNN) 
have enabled unprecedented progress on a wide array of computer vision tasks.
One disadvantage of these approaches is their resource consumption: 
Training deep models within a reasonable amount of time requires special
Graphical Processing Units (GPU) with numerous cores and large memory capacity.
Given the practical importance of these models, a lot of research effort has been directed
towards algorithmic and hardware innovations to improve their resource efficiency such as low-precision arithmetic \cite{jacob2018quantization}, network pruning for inference \cite{molchanov2016pruning}, or efficient stochastic optimization algorithms \cite{kingma2014adam}.

% 2. Benefits of low-memory training
In this paper, we focus on a particular aspect of resource efficiency: 
optimizing the memory cost of training CNNs. 
Given the ubiquity of CNN for practical computer vision applications,
optimizing the memory consumption of CNN training has the potential to
impact a wide range of applications. 
Here, we only present a few of the most interesting 
potential impacts of such optimization:

@@

% 2.3
\textbf{Research in optimization:}
Recent works on stochastic optimization algorithms have highlighted the benefits of large batch training \cite{shallue2018measuring,largebatch,xxx,xxx}.% Add lanmb and lars references
For example, in Imagenet, linear speed-ups in training have been observed 
with increasing batch sizes up to tens of thousands of samples \cite{largebatch}.
Optimizing the memory cost of CNN training may allow further research 
on the optimization trade-offs of large batch training.
Very large batch training on small datasets like MNIST and CIFAR10 is 
computationally inefficient with current stochastic optimization algorithms \cite{largebatch}.
However, for such small datasets, memory optimization would allow to process the full dataset 
in one pass through the networks. 
The ability to process the full dataset in one pass allows to easily train CNNs on the true gradient of the error.
Hence, memory optimization techniques opens the door for research on gradient descent 
optimization of neural networks outside the realm of Stochastic Gradient Descent.

% Paper Overview
% Inherent trade off
There is an inherent trade-off between the memory consumption and computation wall time
of the CNN training procedure:
Previous approaches in optimizing the memory consumption of training CNN,
including gradient checkpointing and reversible network architectures,
trade off memory consumption for additional computations by 
recomputing all or a subset of the hidden layers activations during the backward pass.

% Our approach
Instead, our approach reduce the memory consumption without introducing any additional
computation by leveraging an under-utilized resource: the CPU memory.
We propose to temporarily offloading GPU memory buffers to the CPU during the forward pass 
of the computation, and transfering these memory buffers back into GPU memory as needed by the 
backward pass of the backpropagation algorithm to compute the gradients.

% Key challege
They key challenge in our approach is to efficiently overlap the computation
and data transfers between CPU and GPU to minimize the overhead in wall time
introduced by these data transfer.
We describe an efficient implementation of this approach that allows us to reduce by
up to XXX \% the memory cost of training a VGG network with a minimal wall time overhead of XXX.
We compare the memory vs. wall time trade off of our approach to gradient checkpointing 
to illustrate the efficiency of our approach.

% Summary
The remainer of this paper is organized as follows: In Section 2, we briefly review related work.
Section 3 introduces the preliminary notions necessary to understand the root of the GPU memory bottleneck.
Our approach is described in detail in Section 4, and Section 5 presents the results of our evaluation.

@@

% Pruning & lottery ticket
Network pruning \cite{molchanov2016pruning} is a set of techniques 
developed to decrease the model weight size and computational complexity.
Network pruning works by removing the network weights that contribute the least to the model output.
Pruning deep models has been shown to drastically reduce the memory 
cost and computational cost of inference without  significantly hurting model accuracy.
Although pruning has been concerned with optimization of the resource inference, 
the recently proposed lottery ticket hypothesis \cite{frankle2018lottery} 
has shown that specifically pruned networks could  be trained from scratch to high accuracy.
This may be an interesting and complementary line of work to investigate in the future to reduce training memory costs.

@@

Reversible models \cite{xxx} constrain the CNN architecture to feature invertible transformations.
This allows the activation values of lower layers to be reconstructed from those of higher layers during
the backward pass. 
Reversible networks have been shown to offer a better memory/computation trade-off than 
gradient checkpointing at the cost of constraining the CNN architecture.

Our best performing model combines revertible operations with CPU offloading:
we use the invertible BN-Leaky ReLu block design proposed in \cite{xxx} 
to efficiently deal with normalization and non-linearity layers,
and offload to CPU the activations of the pooling and convolution layers.

@@

\begin{equation}
\frac{\delta \mathcal{L}}{\delta z_{i-1}^j}  = \begin{cases}
\frac{\delta \mathcal{L}}{\delta z_{i-1}^j}, & \text{if}\ z_{i-1}^j \geq 0 \\
0 & \text{if}\ z_{i-1}^j < 0
\end{cases} \\
%\frac{\delta \mathcal{L}}{\delta z_i} \times xxx
\end{equation}

@@

% Describe the ides
The input activations of each layer are kept in GPU memory only to be used for the
computation of the layer weight gradients during the backward pass. 
Hence, the activations of lower layers are kept idle in GPU memory during the
full time span of the forward and backward computations through higher layers.
We propose to offload these activations to the CPU during this idle time
in order to free up some memory space for the computation of higher layers activations.

% Illustrate with figure
Figure XXX illustrates our approach. 
During the forward pass (top), activations are computed forward through the network layers.
Instead of keeping these activations idle in GPU memory, activation values are transfered to 
the CPU memory immediatly after their computation.
In the backward pass (bottom), gradients are backpropagated backward 
through the networ layers following equations XXX.
Our implementation synchronizes the transfer of the layers input activations back to GPU
right before they are needed for their layer's gradient computation.
Hence the key challenge in our implementation consists in synchronizing the data
transfers with the computations so that only the minimal amount of activation values 
is loaded in GPU memory at any given time, while the least amount of time is spent 
waiting for the data transfer.

% Overview of optimizations
To achieve this goal, we propoze optimization along two axes:
The first consists in optimizing the data transfer speed between CPU and GPU memory,
using efficient memory access and data compression schemes.
The second consists in efficient parallelization to maximally overlap the computations
with the data transfer.
The following subsections details optimizations along these two axes.
% For example, when the forward pass through the first layer completes,  
% our algorithm offloads the input activation of
% first layer to CPU memory and forward its output to the
% second layer. In the backward pass, the input data of each
% layer is transferred back to GPU memory right before
% computation of the weight gradients.

@@

% Idea
Figure XXX illustrates the execution through time of a forward 
and backward pass through a toy network with and without parallelization of the data transfer.
Without parallelization, computation and data transfers are performed sequentially 
so that the total wall time is given by the sum of the computation and data transfer time.
$\mathcal{T}_{total} = \mathcal{T}_{comp} + \mathcal{T}_{data}$
Parallelization aims to overlap the computation and data transfer so that the total wall time is given by
$\mathcal{T}_{total} = \mathcal{T}_{comp} + \mathcal{T}_{idle}$, where $\mathcal{T}_{idle}$ represents synchronization delays in cases where the computation is stopped to await for the required data transfer to complete.

The key challenge in this parallelization scheme consists in efficiently
managaing memory allocation and data transfer so as to
minimize the time $\mathcal{T}_{idle}$ spent awaiting for data transfer.
In this paper, we adopt a simple parallelization strategy:
During the forward pass, activations $z_i$ are transfered to CPU 
upon computation by layer $i$ as $z_i = f_i(z_{i-1})$,
and the GPU memory buffered are freed as the next layer computation 
completes as $z_{i+1} = f_{i+1}(z_{i})$.

\begin{algorithm}[h]
	\KwData{Layer $f_i$, input activation $z_{i-1}$ 
		      \\ \hspace{1.1cm} CPU pinned memory buffer $P_{i-1}$
		      \\ \hspace{1.1cm} CPU thread $T_{data}$
		      \\ \hspace{1.1cm} CUDA events $E_{data}^i$, $E_{comp}^i$}
	          \\ \hspace{1.1cm} CUDA Streams $S_{data}$, $S_{comp}$\\
\vspace{1mm}
	\KwResult{$z_i$}
\vspace{1mm}
    \textit{Allocate}($z_i$)\;
	$S_{comp} \Leftarrow z_i \leftarrow f_i(z_{i-1})$\;
	$S_{comp} \Leftarrow E_{comp}^i$\;
\vspace{1mm}
	\textbf{In Thread}  $T_{data}$:\\
		\hspace{1.1cm} 	$S_{data} \Leftarrow P_{i-1} \leftarrow z_{i-1}$\;
		\hspace{1.1cm}  $S_{data} \Leftarrow E_{data}^i$\;
		\hspace{1.1cm}  \textit{Wait}($E_{data}^i$, $E_{comp}^i$)\;
		\hspace{1.1cm}  \textit{Free}($z_{i-1}$)\;
	\caption{Forward procedure through layer $i$ with parallel CPU offloading. 
				 Double arrows indicate the asynchronous execution of a CUDA directive within a stream.
			     Data transfers are executed within dedicated CUDA stream and CPU thread to
		         synchronize the memory deallocation without blocking the execution of upward layers}
\end{algorithm}

One important exception to this rule concerns skip connections, as illustrated in Figure XXX.
Skip connections induce a delay in the GPU buffer dealocation as the input to residual blocks must
be kept in memory until the end of the residual block computation.
In the experiment section, we will show that this delay slightly degrades the memory consumption 
of CPU offloading for residual networks compared to sequential architectures like VGG.

During the backward pass, the input activation $z_{i-1}$ to layer $i$ must be transferred back into GPU memory
before the backward gradient computations to avoid idle time in the GPU.
Hence, we synchronize the data transfer of $z_{i-1}$ with the beginning of the backpropagation through 
the upward layer $f_{i+1}$.

\begin{algorithm}[h]
	\KwData{Layer $f_i$, input activation $z_{i-1}$ 
		\\ \hspace{1.1cm} CPU pinned memory buffer $P_{i-1}$
		\\ \hspace{1.1cm} CUDA events $E_{data}^i$, $E_{data}^{i+1}$}
     	\\ \hspace{1.1cm} CUDA Streams $S_{data}$, $S_{comp}$%\\
	\vspace{1mm} \\
	\KwResult{$\frac{\delta \mathcal{L}}{\delta z_{i-1}}$, $\frac{\delta \mathcal{L}}{\delta \theta_i}$}
	\vspace{1mm}
	\textit{Free}($z_{i}$)\;
	\textit{Allocate}($z_{i-1}$)\;
	\vspace{1mm}
	$S_{data} \Leftarrow z_{i-1} \leftarrow P_{i-1}$\;
	$S_{data} \Leftarrow E_{data}^i$\;
	\vspace{1mm}
	\textit{Wait}($E_{data}^{i+1}$)\;
	\textit{Allocate}($\frac{\delta \mathcal{L}}{\delta z_{i-1}}$, $\frac{\delta \mathcal{L}}{\delta \theta_i}$)\;
	\vspace{1mm}
	$S_{comp} \Leftarrow \frac{\delta \mathcal{L}}{\delta z_{i-1}} \leftarrow \frac{\delta \mathcal{L}}{\delta z_{i}}  \times \frac{\delta z_{i}}{\delta z_{i-1}}$\;
	$S_{comp} \Leftarrow \frac{\delta \mathcal{L}}{\delta \theta_i} \leftarrow \frac{\delta \mathcal{L}}{\delta z_{i}}  \times \frac{\delta z_{i}}{\delta \theta_i}$\;
	\caption{Backward procedure through layer $i$ with parallel CPU offloading.}
\end{algorithm}

We use threading and locks to handle the synchronization on the GPU side 
and CUDA streams to handle the parallelization on the GPU side. 
Algorithm 1 and 2 provide pseudo-code for the forward and backward pass respectively.

% Discussion on compute time vs. transfer time.
With the above parallelization scheme, the overhead $\mathcal{T}_{idle}$ in wall time 
is given by the sum of the difference in the computation and data transfer time at each layer.
If data transfer is faster than the computation in each layer, 
CPU offloading will reduce memory consumption without any overhead.
On a typical workstation with a single NVIDIA GTX 1080 Ti and PCIe lanes,
we found that to be the roughly the case for convolution layers.

% Batch norm and activations.
However, batch normalization layers and activation layers come with negligible computational cost 
but significant data transfer time so that offloading the input activations of such layer comes with
a significant overhead.
Instead of offloading these activations, we reconstruct their input activations 
from their output using their inverse operation as proposed in \cite{xxx},
and only offload the input activations of the convolution and pooling layers to CPU.

 \subsection{Data Transfer Optimization}
% Intro
The time needed to transfer data between devices is given by the product of the
transfer speed bandwidth $\mathcal{B}$ in bits per second by the volume 
$\mathcal{V}$ of the data buffer in bits:

@@

In this section, we propose to minimize data transfer time by simultaneously 
maximizing the transfer speed $\mathcal{B}$
using pinned CPU memory buffers and 
minimizing the activation volume $\mathcal{V}$ by compressing data before transfer.

% Pinned Memory
CPU memory can be accessed through either pageable memory or pinned memory address spaces.
Pinned memory allows for Direct Memory Access (DMA), 
which can significantly speedup data transfer between CPU and GPU devices through PCIe lanes.
However, allocating pinned memory buffers is a time consuming operation, as illustrated in Table XXX.
Hence naively allocating pinned memory buffers within each forward pass would
result in slow data transfer rate and slow down training, which is undesirable.
Instead we propose to allocate the pinned memory buffers once at the initialization of the CNN.
Our implementation scans the network architecture upon instantiation
and allocates dedicated pinned memory buffers for each 
target layer activations (convolutions and max-pooling). 
During the forward pass, we transfer the GPU data directly into the pre-allocated pinned memory buffer,
and, during the backward pass, we transfer the activations back to GPU from these pinned buffers.
Table XXX quantifies the gains in transfer speed brought by the use of pinned memory buffers.

\begin{table*}[t]
\begin{center}
\begin{tabular}{ c c }	
\hline
Operation     & Speed (MB/s)  \\
\hline
Pinned Memory Allocation     		      &  xxx \\
GPU $\rightarrow$ CPU (Pageable)	&  xxx \\
GPU $\rightarrow$ CPU (Pinned)	     &  xxx \\
CPU $\rightarrow$ GPU (Pageable)	&  xxx \\
CPU $\rightarrow$ GPU (Pinned)	     &  xxx \\
\hline
\end{tabular}
\caption{xxx}
\end{center}
\end{table*}

% Compression
Another way to reduce the time of data transfer between CPU and GPU is to compress
the hidden layer activations before transferring them between devices.
Compression can be achieved by diverse means including sparsification, 
quantization or low rank factorization techniques.
Low-rank factorization would induce non-negligible additional computational costs, 
which would undesirably slow down the computation.
Sparsifying the layers activations is an interesting direction, but the affect of sparsification on 
training dynamics is poorly understood so we .
A number of work have shown that neural networks can train 
to equally high accuracy with low precision arithmetic.
In the experiment section, we present results obtained by reducing input activations to half precision,
which allows us to halve the data transfer time without any loss in accuracy,
although more agressive quantization strategies would be interesting to investigate in future work.

>>>>

\usepackage{amsmath}
 \usepackage{titling}
 \usepackage[ruled, vlined]{algorithm2e}
 \usepackage{graphicx}

@@

In recent years, Convolutional Neural Networks (CNN) have enabled 
unprecedented progress on a wide range of computer vision tasks. 
However, training large CNNs is a resource intensive task that requires 
specialized Graphical Processing Units (GPU) and highly optimized 
implementations to get optimal performance from the hardware. 
%Given the ubiquity of CNN for practical computer vision applications,
%considerable efforts have been undertaken to optimize their resource consumption.
GPU memory is a major bottleneck of the CNN training procedure,
 limiting  the size of both inputs and model architectures.
%The backpropagation algorithm requires to accumulate the activation 
%values of hidden layers in live GPU memory during the forward pass
%as these activations are needed for the computation of the weights 
%gradient during the backward pass of the backpropagation algorithm.
In this paper, we propose to alleviate this memory bottleneck by 
leveraging an under-utilized resource of modern systems: 
the device to host bandwidth.
%offloading the input activation of hidden layers to the CPU:
Our method, termed CPU offloading, works by transfering hidden activations to the CPU upon
computation, in order to free GPU memory for upstream layer computations during the forward pass. 
These activations are then transfered back to the GPU as needed by the gradient computations of the backward pass.
The key challenge to our method is to efficiently overlap data transfers and computations
in order to minimize  wall time overhead induced by the additional data transfers.
On a typical work station with a Nvidia Titan X GPU and XXX PCIe lanes, 
we show that our method compares favorably to 
gradient checkpointing as we are able reduce the memory consumption of traning a VGG19 model by 35\% with a minimal additional wall time overhead of 21\%. 
Further experiments detail the impact of the different optimization tricks we propose.
Our method is orthogonal to other techniques for memory reduction such as 
quantization and sparsification so that they can easily be combined for further optimizations.

@@

\section{Introduction}
% 1. Catchy Intro
Over the last few years, Convolutional Neural Networks (CNN) 
have enabled unprecedented progress on a wide array of computer vision tasks.
One disadvantage of these approaches is their resource consumption: 
Training deep models within a reasonable amount of time requires special
Graphical Processing Units (GPU) with numerous cores and large memory capacity.
Given the practical importance of these models, a lot of research effort has been directed
towards algorithmic and hardware innovations to improve their resource efficiency such as low-precision arithmetic \cite{jacob2018quantization}, network pruning \cite{molchanov2016pruning}, or efficient stochastic optimization algorithms \cite{kingma2014adam}.

% 2. Benefits of low-memory training
In this paper, we focus on a particular aspect of resource efficiency: 
optimizing the GPU memory cost of training CNNs. 
Given the ubiquity of CNN for practical computer vision applications,
optimizing the memory consumption of CNN training has the potential to
impact a wide range of applications. 
Here, we only present a few of the most interesting 
potential impacts of such optimization:

@@

% 2.3
\textbf{Research in optimization:}
Recent works on stochastic optimization algorithms have highlighted the benefits of large batch training \cite{shallue2018measuring,mccandlish2018empirical,you2017large}.
For example, in Imagenet, linear speed-ups in training have been observed 
with increasing batch sizes up to tens of thousands of samples \cite{mccandlish2018empirical}.
Optimizing the memory cost of CNN training may allow further research 
on the optimization trade-offs of large batch training.
Very large batch training on small datasets like MNIST and CIFAR10 is 
computationally inefficient with current stochastic optimization algorithms \cite{mccandlish2018empirical}.
However, for such small datasets, memory optimization would allow to process the full dataset 
in one pass through the networks. 
The ability to process the full dataset in one pass allows to easily train CNNs on the exact gradient of the loss function.
Hence, memory optimization techniques opens the door for research on gradient descent 
optimization of neural networks outside the realm of Stochastic Gradient Descent.

% Paper Overview
% Inherent trade off
There is an inherent trade-off between the memory consumption and computation wall time
of the CNN training procedure:
Existing approaches to optimize the memory consumption of CNN training,
(gradient checkpointing, reversible network architectures),
trade off memory consumption for additional computations by 
recomputing all or a subset of the hidden layers activations during the backward pass.

% Our approach
Instead, our approach reduces the GPU memory consumption without introducing any additional
computation by leveraging an under-utilized resource: host-device communication.
We propose to temporarily offload GPU memory buffers to the CPU during the forward pass 
of the computation, and transfering these memory buffers back into GPU memory as needed by the gradient computations during the backward of the backpropagation algorithm.

% Key challege
They key challenge in our approach is to efficiently overlap the GPU computations
with the data transfers between CPU and GPU in order to minimize the overhead in wall time
introduced by these data transfer.
In this paper, we describe an efficient implementation of this approach that allows us to reduce by
up to 35\% the memory cost of training a VGG network with a minimal wall time overhead of 21\%.
We compare the memory vs. wall time trade off of our approach to gradient checkpointing 
to illustrate the efficiency of our approach.

% Summary
The remainer of this paper is organized as follows: In Section 2, we briefly review the literature for related work.
Section 3 introduces the preliminary notions necessary to understand the root of the GPU memory bottleneck.
Section 4 presents our approach and details the different tricks needed for efficient implementation. 
Finally, Section 5 presents the results of our evaluation.

@@

% Pruning & lottery ticket
Network pruning \cite{molchanov2016pruning} is a set of techniques 
developed to decrease the model weight size and computational complexity.
Network pruning works by removing the network weights that contribute the least to the model output.
Pruning deep models has been shown to efficiently reduce the memory 
and computational cost of inference without  significantly hurting model accuracy.
Although pruning methods focus on the optimization of inference, 
the recently proposed lottery ticket hypothesis \cite{frankle2018lottery} 
has shown that specifically pruned networks could  be trained from scratch to high accuracy.
This may be an interesting and complementary line of work to investigate in the future to reduce training memory costs.

@@

Reversible models \cite{gomez2017reversible,jacobsen2018revnet} constrain the CNN architecture to feature invertible transformations.
This allows the activation values of lower layers to be reconstructed from those of higher layers during
the backward pass. 
Reversible networks have been shown to offer a better memory/computation trade-off than 
gradient checkpointing at the cost of constraining the CNN architecture.

Our approach combines revertible operations with CPU offloading:
we use the invertible BN-Leaky ReLu block design proposed in \cite{rota2018place} 
to efficiently deal with normalization and non-linearity layers,
and only offload to CPU the activations of the pooling and convolution layers.

@@

\begin{equation}
\frac{\delta \mathcal{L}}{\delta z_{i-1}^j}  = \begin{cases}
\frac{\delta \mathcal{L}}{\delta z_{i-1}^j}, & \text{if}\ z_{i-1}^j \geq 0 \\
0 & \text{if}\ z_{i-1}^j < 0
\end{cases} \\
\end{equation}

@@

% Describe the ides
The input activations of each layer are kept in GPU memory only to be used for the
computation of the layer weight gradients during the backward pass. 
Hence, the activations of lower layers are kept idle in GPU memory during the
forward and backward computations through higher layers.
We propose to offload these activations to the CPU during this idle time
in order to free up some GPU memory space for the computation of higher layers activations.

% Illustrate with figure
Figure 1 illustrates our approach. 
During the forward pass (top), activations are computed forward through the network layers.
Instead of keeping these activations idle in GPU memory, activation values are transfered to 
the CPU memory immediately after their computation.
In the backward pass (bottom), gradients are backpropagated backward 
through the network layers following equations (3).
Our implementation synchronizes the transfer of the layers input activations back to GPU
right before they are needed for their layer's gradient computation.
Hence the key challenge in our implementation consists in synchronizing the data
transfers with the computations so that only the minimal amount of activation values 
is loaded in GPU memory at any given time, while the least amount of time is spent 
waiting for the data transfer.

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{Figure1.jpg}
\caption{Illustration of CPU Oflloading. 
During the forward pass (top), 
hidden activation buffers are transfered to CPU memory upon computation.
During the backward pass (bottom), hidden activation buffers 
are transfered back to GPU memory just in time to compute the 
weight gradients.}
\end{figure}

% Overview of optimizations
To achieve this goal, we propose a set of optimization tricks organized along two axes:
The first consists in optimizing the data transfer speed between CPU and GPU memory,
using efficient memory accesses and data compression schemes.
The second consists in efficient parallelization to maximally overlap the computations
with the data transfer.
The following subsections details optimizations along these two axes.
% For example, when the forward pass through the first layer completes,  
% our algorithm offloads the input activation of
% first layer to CPU memory and forward its output to the
% second layer. In the backward pass, the input data of each
% layer is transferred back to GPU memory right before
% computation of the weight gradients.

@@

% Idea
Figure 2 illustrates the execution through time of a forward 
and backward pass through a toy network with and without parallelization of the data transfer.
Without parallelization, computation and data transfers are performed sequentially 
so that the total wall time is given by the sum of the computation and data transfer time.
$\mathcal{T}_{total} = \mathcal{T}_{comp} + \mathcal{T}_{data}$
Parallelization aims to overlap the computation and data transfer so that the total wall time is given by
$\mathcal{T}_{total} = \mathcal{T}_{comp} + \mathcal{T}_{idle}$, where $\mathcal{T}_{idle}$ represents synchronization delays in cases where the computation is stopped to await for the required data transfer to complete.

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{Figure2.jpg}
\caption{Illustration of the parallel CPU Offloading execution through time.
(top): Computations and data transfer are executed in parallel in their dedicated stream. 
The overhead in wall time $\mathcal{T}_{idle}$ is spent as the computation stream awaits 
for data transfers to complete (illustrated in red). 
(bottom): Without parallelism, computations and data transfers are sequentially executed within the same stream
so that the overhead in wall time corresponds to the total time of data transfers $\mathcal{T}_{data}$}.
\end{figure}

The key challenge in this parallelization scheme consists in efficiently
managaing memory allocation and data transfer so as to
minimize the time $\mathcal{T}_{idle}$ spent awaiting for data transfer.
In this paper, we adopt a simple parallelization strategy:
During the forward pass, activations $z_i$ are transfered to CPU 
upon computation by layer $i$ as $z_i = f_i(z_{i-1})$,
The GPU memory buffer of $z_i$ must then await for both the data transfer to CPU and
the next layer computation $z_{i+1} = f_{i+1}(z_{i})$ to complete before deallocation in order to
avoid unfortunate overwriting by concurrent operations.

\begin{figure*}[h]
\begin{algorithm}[H]
	\KwData{Layer $f_i$, input activation $z_{i-1}$ 
		      \\ \hspace{1.1cm} CPU pinned memory buffer $P_{i-1}$
		      \\ \hspace{1.1cm} CPU thread $T_{data}$
		      \\ \hspace{1.1cm} CUDA events $E_{data}^i$, $E_{comp}^i$
	          \\ \hspace{1.1cm} CUDA Streams $S_{data}$, $S_{comp}$}
\vspace{1mm}
	\KwResult{$z_i$}
\vspace{1mm}
    \textit{Allocate}($z_i$)\;
	$S_{comp} \Leftarrow z_i \leftarrow f_i(z_{i-1})$\;
	$S_{comp} \Leftarrow E_{comp}^i$\;
\vspace{1mm}
	\textbf{In Thread}  $T_{data}$:\\
		\hspace{1.1cm} 	$S_{data} \Leftarrow P_{i-1} \leftarrow z_{i-1}$\;
		\hspace{1.1cm}  $S_{data} \Leftarrow E_{data}^i$\;
		\hspace{1.1cm}  \textit{Wait}($E_{data}^i$, $E_{comp}^i$)\;
		\hspace{1.1cm}  \textit{Free}($z_{i-1}$)\;
	\caption{Forward procedure through layer $i$ with parallel CPU offloading. }
\end{algorithm}
\small{Algorithm 1: Double arrows indicate the asynchronous execution of a CUDA directive within a stream.
Data transfers are executed within dedicated CUDA stream and CPU thread to
synchronize the memory deallocation without blocking the execution of upward layers}
\end{figure*}

One important exception to this rule concerns skip connections, as illustrated in Figure 3.
Skip connections induce a delay in the GPU buffer dealocation as the input of residual blocks must
be kept in memory until the end of the residual block computation to be added to the output.
In ResNet architectures, this delay is short enough to have little impact on the memory/wall time trade off.
However, this means that our method is poorly suited to densely connected 
networks such as DenseNet \cite{huang2017densely} or UNet \cite{ronneberger2015u}, which features very long skip connections.

%() In the experiment section, we will show that this delay slightly degrades the memory consumption 
%of CPU offloading for residual networks compared to sequential architectures like VGG.
%Synchronization of the computation and data transfers during the backward pass is slightly more complicated:

During the backward pass, the input activation $z_{i-1}$ 
of layer $i$ must be transferred back into GPU memory
for the backward gradient computations to proceed.
Hence, we overlap the transfer of $z_{i-1}$ to GPU 
with the backward computations of the upper layer $i+1$ to avoid GPU idle time.
As formalized in Algorithm 2, we thus synchronize the data transfer of $z_{i-1}$ 
with the beginning of the backpropagation through $f_{i+1}$.

\begin{algorithm}[h]
	\KwData{Layer $f_i$, output gradients $\frac{\delta \mathcal{L}}{\delta z_{i}}$
		\\ \hspace{1.1cm} CPU pinned memory buffer $P_{i-1}$
		\\ \hspace{1.1cm} CPU thread $T_{comp}$
		\\ \hspace{1.1cm} CUDA events $E_{data}^i$, $E_{data}^{i+1}$}, $E_{comp}^{i}$
     	\\ \hspace{1.1cm} CUDA Streams $S_{data}$, $S_{comp}$%\\
	\vspace{1mm} \\
	\KwResult{$\frac{\delta \mathcal{L}}{\delta z_{i-1}}$, $\frac{\delta \mathcal{L}}{\delta \theta_i}$}
	\vspace{1mm}
	%\textit{Free}($z_{i}$)\;
	\textit{Allocate}($z_{i-1}$)\;
	\vspace{1mm}
	$S_{data} \Leftarrow z_{i-1} \leftarrow P_{i-1}$\;
	$S_{data} \Leftarrow E_{data}^i$\;
	\vspace{1mm}
	\textit{Wait}($E_{data}^{i+1}$)\;
	\textit{Allocate}($\frac{\delta \mathcal{L}}{\delta z_{i-1}}$, $\frac{\delta \mathcal{L}}{\delta \theta_i}$)\;
	\vspace{1mm}
	$S_{comp} \Leftarrow \frac{\delta \mathcal{L}}{\delta z_{i-1}} \leftarrow \frac{\delta \mathcal{L}}{\delta z_{i}}  \times \frac{\delta z_{i}}{\delta z_{i-1}}$\;
	$S_{comp} \Leftarrow \frac{\delta \mathcal{L}}{\delta \theta_i} \leftarrow \frac{\delta \mathcal{L}}{\delta z_{i}}  \times \frac{\delta z_{i}}{\delta \theta_i}$\;
	\caption{Backward procedure through layer $i$ with parallel CPU offloading.}
\end{algorithm}

We use threading and locks to handle the parallelism and synchronization on the CPU side,
and CUDA streams to handle the parallelization on the GPU end. 
CUDA events provides synchronization primitives to precisdely track 
the completion of kernel executions.
Algorithm 1 and 2 provide pseudo-code for the forward and backward pass respectively.

\begin{figure}[h]
\includegraphics[width=\textwidth]{Figure3.png}
\caption{Illustration of the delay in memory deallocation induced by residual connections.
(left) without residual connection, the input $z_{i-1}$ of layer $f_i$ can be freed from GPU memory 
as $f_i$ computation terminates. 
(right) Residual connections induce a delay in memory deallocation
as $z_{i-1}$ must be kept in GPU memory to be added to the ouput of the residual block. 
}
\end{figure}

% Discussion on compute time vs. transfer time.
With the above parallelization scheme, the overhead $\mathcal{T}_{idle}$ in wall time 
is given by the sum of the difference in the computation and data transfer time at each layer.
If data transfer is faster than the computation in each layer, 
CPU offloading will reduce memory consumption without any overhead in wall time.
Hence, CPU offloading is best suited for computation intensive operations like convolution.
In contrast, batch normalization layers come with negligible computational cost,
but significant data transfer time so that offloading the input activations of 
such layer comes with a significant overhead.
Instead of offloading these layers, we reconstruct their input activations 
from their output using their inverse operation, as proposed in \cite{rota2018place},
and only offload the input activations of the convolution and pooling layers to CPU.

 \subsection{Data Transfer Optimization}
% Intro
The time needed to transfer data between devices is given by the product of the
transfer speed bandwidth $\mathcal{B}$, in bytes per second, by the volume 
$\mathcal{V}$ of the data buffer in bytes:

@@

Hence, minimizing data transfer time can be achieved by either
maximizing the transfer speed $\mathcal{B}$ or minimizing
the activation volume $\mathcal{V}$ through data compression.
In this paper, we focus on maximizing data transfer speed using pinned
CPU memory buffers.

% Pinned Memory
CPU memory can be accessed through either pageable memory 
or pinned memory address spaces.
Pinned memory allows for Direct Memory Access (DMA), 
which can significantly speedup data transfer between CPU and GPU devices through PCIe lanes.
However, allocating pinned memory buffers is a time consuming operation, as illustrated in Table 1.
Hence, naively allocating pinned memory buffers within each forward pass would
slow down training due to the overhead of pinned memory allocation.
Instead, we propose to allocate the pinned memory buffers beforehand, 
during the initialization of the CNN.
Our implementation scans the network architecture upon instantiation
and allocates dedicated pinned memory buffers for each 
target layer activations (convolutions and max-pooling). 
During the forward pass, we transfer the GPU data directly into the pre-allocated pinned memory buffer,
and, during the backward pass, we transfer the activations back to GPU from these pinned buffers.
Table 1 quantifies the gains in transfer speed brought by the use of pinned memory buffers, and in Section5, we evaluate the impact of using pinned memory buffers on the overall efficiency of our method.

\begin{table*}[t]
\begin{center}
\begin{tabular}{ c c }	
\hline
Operation     & Speed (GB/s)  \\
\hline
Pinned Memory Allocation     		&  $1.95$        \\
GPU $\rightarrow$ CPU (Pageable)	&  $1.75$        \\
GPU $\rightarrow$ CPU (Pinned)	        &  $6.5$          \\
CPU $\rightarrow$ GPU (Pageable)	&  $4.1$          \\
CPU $\rightarrow$ GPU (Pinned)	        &  $6.1$          \\
\hline
\end{tabular}
\caption{Measurement of data transfer and pinned memory allocation speed on our workstations.
Pinned memory buffers bring a substantial speed up in both directions of the data transfer.}
\end{center}
\end{table*}

% Compression
Another way to reduce the time of data transfer between CPU and GPU is to minize 
$\mathcal{V}$ by compressing the hidden layer activations 
before transferring them between devices.
Compression can be achieved by diverse means including sparsification, 
quantization or low rank factorization techniques.
Low-rank factorization would induce non-negligible additional computational costs, 
which would slow down the computation.
Either sparsifying or quantizing the layers activations are interesting solutions.
However, compression would incur minor losses of precision in the activation values,
which might affect the training dynamics.
Although previous works suggest that such minor losses in precision does not negatively impact training,
investigation of compression techniques would require a detailed analysis 
of the impact of compression on accuracy, which is out of the scope of this paper.
Hence we leave this question open for future work as an interesting direction for further optimization.
